---
title: "Lab 2: Grammar of graphics"
subtitle: "MATH1115 -- Interrogating Data"
author: "&copy; University of Sydney `r format(Sys.time(), '%Y')`"
date: 
output:
  html_document:
    code_download: true # embed .Rmd file
    toc: true  # table of contents
    toc_depth: 3
    toc_float: true # toc is always visible when scrolled
    number_sections: true 
    df_print: paged # format data frame outputs automatically
    theme: default # theme the doc
    highlight: textmate # theme syntax colouring
---

<!-- ######################################### -->
<!-- #### Style settings - do not delete! #### -->

```{css newstyles, echo=FALSE}
body{ /* Most readable font (esp < 8): */
  font-family: Verdana;
}
h1, .h1, h2, .h2, h3, .h3 { /* Add space before headings: */
    margin-top: 56px;
}
h1, h2 { /* add border to h1 and h2 */
  border-bottom: solid 1px #666;
}
h2 { /* Resize header 2: */
  font-size: 22px;
}
body { /* Make main text colour black */
  color: black;
}
a { /* Link colours */
  color: blue;
}
.tocify { /* Some toc fixes*/
  width: 100% !important;
  border: none; /* remove border */
}
/* Space Between TOC and 
Righthand side content on large screens */

@media (min-width: 992px) {
  .col-md-9 {
    width: 75%;
    padding-left: 5em !important;
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- #########  End style settings. ########## -->
<!-- ######################################### -->


# The ggplot2

- Currently part of the tidyverse, allows us to compose graphs by combining *layers* of information, in a manner that can be structured and understood in plain language. This is achievable because `ggplot2` has an underlying grammar structure based on the [Grammar of Graphics](https://vita.had.co.nz/papers/layered-grammar.html).

- The basic components of a `ggplot()` are:
    1. The **canvas**, where you map your aesthetics
    2. The **geometric layer(s)**, where you generate the plot
    3. The details, which involve **scales, axes, legends and facets**

The basic grammar of a `ggplot()` relies on adding "blocks" on top of one another with the help of the `+` operator:

```
canvas +
  geometric layer +
  scale +
  legends +
  facets +
  ...
```

## Tidyverse

Load the `tidyverse` package using `library()`. If you need to install it, do it **in the console**. We have deliberately commented the install code so that you do not run them in a code chunk!

```{r, message=FALSE}
# install.packages("tidyverse")
library(tidyverse)  # loads the packages we need: dplyr and ggplot2
```

# Demo

## The canvas

What is the canvas? What are aesthetics? Let's view a basic `ggplot()` output where we do not add a geometric layer, using the `airquality` dataset that you are now familiar with from the first lab.

```{r}
# canvas only
ggplot(data = airquality)
```

Even when a dataset is provided, only a blank canvas is produced unless we provide additional information in the form of **aesthetics**. Let's define some using `aes()`.

```{r}
# canvas and aesthetics
ggplot(airquality, aes(x = Temp))
```

What has changed? Try to change the variable in `x` to `Solar.R` or `Temp`. What happens to the x-axis?

- It gives different range of the x-axis, based on the dataset given.


## Geometric layer(s)

In the first lab, we looked at the histogram of `Ozone` from the `airquality` dataset, using base R graphics:

```{r}
# histogram of `Ozone` in `airquality` using base R plot
hist(airquality$Ozone, 
  col = "lightgreen", 
  xlab = "Mean Ozone (ppb)", 
  main = "New York Air Quality (May-Sept 1973)")
```

Let's create a histogram of the same data using `ggplot()`. We do this by adding a `geom_histogram()` layer after the canvas, using the `+` operator.

```{r, message=FALSE, warning=FALSE}
# histogram of `Ozone` in `airquality` using ggplot2
ggplot(airquality, aes(x = Ozone)) +  # create canvas + aes, then
  geom_histogram() # generate a histogram
```

- Notice the message about `stat_bin()` and non-finite values. Why do you think they appear? 
  - The stat_bin is because the default-30-bar graph set can be better
  - And there are blank data slot, as non-finite values.
  
- What else can we do to make the plot look more like the one made in base R? (We can come back to this question later.)

Customize it a bit:
```{r, message=FALSE, warning=FALSE}
ggplot(airquality, aes(x = Ozone)) + # create canvas + aes, then
  # generate a histogram:
  geom_histogram(fill = "lightgreen", # fill bars
                 colour = "black",    # colour lines
                 bins = 9) +          # no. of bars
                # Or can use binwidth: this defines how thick the bar is.
  ylab("Frequency")  # rename y axis label
```

Geometric layers, or geoms, are defined by the `geom_*()` function. The `ggplot()` package provides a plethora geoms which can automatically map the data defined in `aes()`, into the plots! Try the following:

- `geom_boxplot()`
- `geom_density()`
- `geom_bar()`
- `geom_dotplot()`

```{r, message=FALSE, warning=FALSE}
# a density plot using ggplot2:
ggplot(airquality, aes(x = Ozone)) +
  geom_density(fill = "lightgreen") 
```

## Multiple layers

We can use different aesthetics, based on the same data, to plot *both* a histogram and a density plot.

```{r, message=FALSE, warning=FALSE}
ggplot(airquality) +  # create canvas, then...
  # add histogram, then...
  geom_histogram(aes(x = Ozone, y = ..density..), 
                 fill = "lightgreen",
                 colour = "black",
                 bins = 9) + # the magic
  # add density plot layer
  geom_density(aes(x = Ozone), 
               size = 2)  
```
Reminder: the order of the layer is back + front + fronter + ...

## Positioning the aesthetics

The plot above is possible because we can define different aesthetics in each geom. Below is another example using bivariate data from `airquality`.

```{r, message=FALSE, warning=FALSE}
# constructing a scatter plot with best-fit regression line
ggplot(airquality, aes(Solar.R, Temp)) +
  geom_point() + 
  geom_smooth(method = "lm") +  # generate linear regression line (lm)
  xlab("Solar radiation (Angstroms)") +  # set x-axis label
  # y-axis label with special character:
  ylab(expression("Temperature " ( degree*C))) 

```

All the code below will produce the exact same plot:
- Because aesthetics can be on any layout, know it's flexible.

```{r, eval=FALSE}

# move aesthetics to the geoms:
ggplot(airquality) +
  geom_point(aes(Solar.R, Temp)) +
  geom_smooth(aes(Solar.R, Temp),
              method = "lm") + 
  xlab("Solar radiation (Angstroms)") +  # set x-axis label
  # y-axis label with special character:
  ylab(expression("Temperature " ( degree*C))) 

# move data canvas and aesthetics to the geoms:
ggplot() +
  geom_point(data = airquality, aes(Solar.R, Temp)) +
  geom_smooth(data = airquality, aes(Solar.R, Temp),
              method = "lm") + 
  xlab("Solar radiation (Angstroms)") +  # set x-axis label
  # y-axis label with special character:
  ylab(expression("Temperature " ( degree*C))) 
```


## Grouping by a third variable (dimensions)

- [Simpson's paradox](https://en.wikipedia.org/wiki/Simpson%27s_paradox)
  - Basically means data cannot be representative by simply adding up to show something, there are other factors can impact on them.
- To understand whether there are differences between categories of data. 

We can use `Month` to group our data by colour in `ggplot2`. However, it is currently a `numeric` variable. Let's add  `Month` as a third aesthetic. 

```{r, message=FALSE, warning=FALSE}
ggplot(airquality, aes(Solar.R, Temp, 
                       colour = Month)) +
  geom_point() +
  geom_smooth(method = "lm") +  
  xlab("Solar radiation (Angstroms)") +  
  ylab(expression("Temperature " ( degree*C))) 
```

Because `Month` has been identified as a continuous variable, the plot has not visualised the data "properly". It is still quite difficult to see differences in the patterns among months.

In the first Lab, we showed that `Month` should be converted to a factor. The reason for doing so is because plots will interpret data differently based on their `class()`.This is the same plot, but with `Month` converted to a variable of class `factor`:

```{r, message=FALSE, warning=FALSE}
# replot the same visualisation, but now with factor(Month)
ggplot(airquality, aes(Solar.R, Temp, 
                       colour = factor(Month))) + # ***
  geom_point() +
  geom_smooth(method = "lm") +  
  xlab("Solar radiation (Angstroms)") +  
  ylab(expression("Temperature " ( degree*C))) 
```

Success! But are we able to look at the plot and easily isolate the raw points?. When using colours or shapes to differentiate groups and the result looks "busy", it is a good idea to **facet** the plot using `facet_wrap()`.

```{r, message=FALSE, warning=FALSE}
# replot the same visualisation, but now with facet_wrap()
ggplot(airquality, aes(Solar.R, Temp, 
                       colour = factor(Month))) + # ***
  geom_point() +
  geom_smooth(method = "lm") +  
  xlab("Solar radiation (Angstroms)") +  
  ylab(expression("Temperature " ( degree*C))) +
  facet_wrap(~ Month)
```


# Explore: boxplot

We have only scratched the surface of what can be done using `ggplot2`.
You will now spend the rest of the lab to explore scales, axes, legends and facets and understand data better.

Create a boxplot using 2 variables: `Ozone` and `Month` with the following conditions:

1. `Ozone` as `y`, and is named appropriately on the axis
2. `Month` as `x`, and is named appropriately on the axis
3. `Month` is renamed to the months May -- Sep, instead of numbers
4. `Month` is arranged by date when plotted
5. All boxplots are filled with the colour `"firebrick"`
6. A theme of your choice is applied to the entire plot.

Hints:

These are real web browser searches. By default, we are using the Google search engine to process these keywords.

- [ggplot rename x axis names](https://www.google.com/search?q=ggplot+rename+x+axis+names)
- [ggplot reorder a variable](https://www.google.com/search?q=ggplot+reorder+a+variable)
- [ggplot change boxplot colour](https://www.google.com/search?q=ggplot+change+boxplot+colour)
- [ggplot add theme](https://www.google.com/search?q=ggplot+add+theme)

```{r}
# create your plot here
```


# Explore: multivariate plot

We will improve on the scatterplot that was produced in the Demo.

```{r, message=FALSE, warning=FALSE}
ggplot(airquality, aes(Solar.R, Temp, colour = factor(Month))) +
  geom_point() +
  geom_smooth(method = "lm") +  
  xlab("Solar radiation (Angstroms)") +  
  ylab(expression("Temperature " ( degree*C))) +
  facet_wrap(~ Month)

```

1. Rename the months from numbers to May -- Sep
2. Rename the Legend title from `"factor(Month)"` to `"Month"`
3. Rearrange the facet so that it is wrapped in two columns, not three columns.
4. Instead of a universal scale, make it such that each plot has its own individual y-axis that is scaled for the data.
5. Move the legend from the right side of the canvas to the top.

Hints:

Search for these keywords yourself:

- ggplot facet_wrap column number
- ggplot facet_wrap individual scale 
- ggplot move legend position top

# Explore: road fatalities

You should have some understanding of how to read `.csv` and `.xlsx` files into R by now.
How do you construct your code such that the `read*()` code works when you transfer it to a group member?

Notes:

- Try to understand how file paths work *relative* to an R Markdown file. 
- If a data file is in the same folder as your `.Rmd` file, then reading the file is as easy as `read.csv("filename.csv")`.
- If the data file is in a folder, and that folder is in the same folder as your `.Rmd` file (e.g. the "data" folder), then you will need to add the folder path to the code: `read.csv("folder/filename.csv")`
- If the file is an Excel file, then you have to use the `readxl` package and read the file using `read_excel()`.


Read the `AllFatalities.csv` file into R and save it as an object, e.g. `fatalities`.

```{r}
# read the .csv file `ardd_fatalities_nov2020.csv`

# use this to check if your data file is in the same directory 
# as this .Rmd file: 
# list.files() 

# use this to determine the working directory of
# this .Rmd file:
# getwd()

# example code:
# fatalities <- read.csv("data/AllFatalities.csv")
```

- Investigate the structure of the data. What variables are interesting?
- Use numerical summaries to find interesting observations.
- Use `ggplot2` to visualise interesting patterns.
- List some research questions that can be answered with the `ggplot2` plots.