---
title: "Review"
author: "Dongyi Guo"
output: rmdformats::robobook
---

<!-- ######################################### -->
<!-- #### Style settings - do not delete! #### -->

```{css newstyles, echo=FALSE}
body{ /* Most readable font (esp < 8): */
  font-family: Verdana;
}
h1, .h1, h2, .h2, h3, .h3 { /* Add space before headings: */
    margin-top: 56px;
}
h1, h2 { /* add border to h1 and h2 */
  border-bottom: solid 1px #666;
}
h2 { /* Resize header 2: */
  font-size: 22px;
}
body { /* Make main text colour black */
  color: black;
}
a { /* Link colours */
  color: blue;
}
.tocify { /* Some toc fixes*/
  width: 100% !important;
  border: none; /* remove border */
}
/* Space Between TOC and 
Righthand side content on large screens */

@media (min-width: 992px) {
  .col-md-9 {
    width: 75%;
    padding-left: 5em !important;
  }
}
```

```{r setup, include=FALSE}
# code chunk settings for entire document
knitr::opts_chunk$set(echo = TRUE,
                      hide = FALSE)
```

<!-- #########  End style settings. ########## -->
<!-- ######################################### -->

# Air quality data

## Domain knowledge

What is the background to the `airquality` dataset?
- ? == man page

```{r}
# in the console, check the documentation for `airquality` using `?`
# ?airquality
```

## Structure

How big is the data? - dim it 
```{r}
dim(airquality)  # check dimensions (row and col)
```

What are the names of the variables? - names the label
```{r}
names(airquality)  # column names if a data frame object is supplied
```

How has R classified the variables? - str to show the type of each labels
```{r}
str(airquality)  # check the structure of `airquality` data
```

Note: The `str()` function can replace `dim()`, `names()`, and `class()`.

* Because it did there jobs...

## Explore 

Investigate the `Ozone` variable. What can you observe?

```{r}
# Isolate variable with $
ozone <- airquality$Ozone
# Numerical summaries
length(ozone)

# This should give length of data under Ozone label
mean(ozone, na.rm = TRUE)  # mean of all values
median(ozone, na.rm = TRUE)  # median of all values
sd(ozone, na.rm = TRUE)  # standard deviation
summary(ozone)  # all summary statistics at once
```
Note: the na.rm means removing all the NA values, if they are involved, mean and median cannot be calculated.
```{r}
# Graphical summaries
hist(ozone)
boxplot(ozone)
```

You can customise and label your data visualisations to create "better" data summaries. 

```{r}
# create a histogram using base R
hist(ozone, 
     col = "lightgreen",  # modify colour fill of bars
     xlab = "Mean Ozone (ppb)",  # x-axis label
     main = "New York Air Quality (May-Sept 1973)")  # add a title

# create a boxplot
boxplot(ozone,
        col = "lightgreen",
        xlab = "Mean Ozone (ppb)",
        main = "New York Air Quality (May - Sept 1973)")
```

Reclassify `Month` as a `factor` and then investigate. What can you observe?

```{r}
## Reclassify variable
class(airquality$Month)
monthF <- as.factor(airquality$Month)
class(monthF)

## view data as a table object
table(monthF)

barplot(table(monthF)) #Can you make this graph look nicer?
```

Filter (or facet) your `ozone` data (quantitative) by `months` (qualitative). What can your observe?

```{r}
boxplot(airquality$Ozone ~ monthF, 
        names = c("May","June","July","Aug","Sept"),
        ylab = "New York Air Quality (May-Sept 1973)")
```

# Import

What is the background to the [BITRE](https://bitre.gov.au/statistics/safety/fatal_road_crash_database.aspx) data:

```{r, eval=TRUE}
library(readxl)
input <- read_excel("data/ardd_fatalities_jul2021.xlsx")
input
```