---
title: "DATA1001/1901 Topic 6"
subtitle: "Understanding Chance"
author: "© University of Sydney 2021"
output:
  html_document:
    code_download: true # embed .Rmd file
    toc: true  # table of contents
    toc_depth: 3
    toc_float: true # toc is always visible when scrolled
    number_sections: true 
    theme: sandstone # theme the doc
    highlight: tango # theme syntax colouring
---



```{css newstyles, echo=FALSE}
h1, .h1, h2, .h2, h3, .h3 { /* Add space before headings: */
    margin-top: 56px;
}
h1, h2 { /* add border to h1 and h2 */
  border-bottom: solid 1px #666;
}
h2 { /* Resize header 2: */
  font-size: 22px;
}
h3 { /* Resize header 2: */
  font-size: 18px;
}
body { /* Make main text colour black */
  color: black;
}
a { /* Link colours */
  color: blue;
}
.tocify { /* Some toc fixes*/
  width: 100% !important;
  border: none; /* remove border */
}
.tocify-header { /* fix for horrible indent in toc */
  text-indent: initial;
}

```

*Note that this document has been updated in 2021 and may look slightly different from the lab videos.*


# Live Demo

We will demonstrate simulation of simple scenarios, including a simple game.

## Simulation

```{r}
set.seed(1)  # This ensures the same simulation, "1" can be whatever you choose

# Simulate 1 fair coin toss
sample(c("H", "T"), 1, replace = TRUE)

# Simulate 1 biased coin toss
sample(c("H", "T"), 1, replace = TRUE, prob = c(0.9, 0.1))

# Simulate 10 fair coin tosses, and summarise chances in a table.

## Option1
sample1 <- sample(c("H", "T"), 10, replace= TRUE)
table(sample1)/10

## Option 2
sample2 <- replicate(10, sample(c("H", "T"), 1, replace = TRUE))
table(sample2) / 10

## Option 3
sample3 <- rbinom(n = 10, 1, 0.5)
table(sample3) / 10
```

## Using binomial function

Suppose we have 10 trials with P(event) = 0.3.

```{r}
# The chance of getting EXACTLY 4 events is:
dbinom(4,10,0.3)  # dbinom(x,n,p)

# The chance of getting LESS OR EQUAL to 5 events is:
pbinom(5,10,0.3)    # pbinom(x,n,p)
```

# Try

## Dice

Here's the rules of a game:

- A game costs $50 to play. 

- You get to roll a fair dice 20 times.

- You get \$8 for any roll that produces a 6.

Simulate your net win/loss.

```{r}
# Simulate the 20 throws of the fair dice
set.seed(100)
sample3 <- sample(c(1:6), 20, replace = TRUE) # why is replace set to TRUE? Can repeat draw

table(sample3) # preview data

# plot
library(ggplot2) # alternative: library(tidyverse)
ggplot() +
  geom_bar(aes(factor(sample3)),# why did we do this? Try removing factor()
           colour = "white", 
           fill = "firebrick") +
  xlab("Rolled number on dice") +
  ylab("Frequency") +
  ggtitle("Simulation of 20 throws of fair dice") + 
  theme_bw() # try theme_minimal() and theme_gray()
  
```


Conclusion: Based on this simulation, your net loss would be $\$8 \times 5 - \$50 = - \$10$.


## Coins

Here's the rules of another game:

- A game costs $10 to play. 
- You get to throw a biased coin 5 times, where the probability of getting a head is 0.2. 
- You get $20 for each Head you throw. 

Simulate your net win/loss.

```{r}
coin <- sample(c(0,1), 5, replace = TRUE, prob = c(0.2,0.8))
```

Do this many times, we can only have 0-2 heads. Thus losing 5 * 10 - (0\~2) * 20 = c(10,50) dollars.
