---
title: "Chi-square and Linear Model"
author: "Dongyi Guo"
output: rmdformats::robobook
---

```{css newstyles, echo=FALSE}
h1, .h1, h2, .h2, h3, .h3 { /* Add space before headings: */
    margin-top: 56px;
}
h1, h2 { /* add border to h1 and h2 */
  border-bottom: solid 1px #666;
}
h2 { /* Resize header 2: */
  font-size: 22px;
}
body { /* Make main text colour black */
  color: black;
}
a { /* Link colours */
  color: blue;
}
.tocify { /* Some toc fixes*/
  width: 100% !important;
  border: none; /* remove border */
}
.tocify-header { /* fix for horrible indent in toc */
  text-indent: initial;
}

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(MASS)
```

# Chi-square Test

## Independence

```{r}
glimpse(Aids2)

chisq.test(table(Aids2$state, Aids2$status)) # chisq.test takes a contingency table of x and y.
```

## Goodness of Fit

```{r}
Aids2 %>% 
  group_by(sex) %>% 
  count -> gender

gender
```

Then use chisq.test(observed count vect, expected probability vect)

```{r}
observed <- gender$n
expected <- c(.5,.5)
chisq.test(observed, p = expected)
```

## homogeneity

```{r}
glimpse(HairEyeColor)
```

Mind this data set puts gender in different groups, select them out:

```{r}
# Select with a_table[row,col,grp]
males <- HairEyeColor[,,"Male"]
females <- HairEyeColor[,,"Female"]
```

Now do chisq.test

```{r}
chisq.test(males, females)
```

There should be warnings, since the chisq.test requires minimum of observation of 5, but we have 3 here. So we need to:

```{r}
chisq.test(males, females, simulate.p.value = TRUE)
```

# Regression Model Testing

We use to use `lm(var1 ~ var2)`.

But this requires var1 and var2 should be completely independent to each other.

Better way to do is:

```{r}
ggplot(iris, aes(Petal.Width, Petal.Length)) +
  geom_point()

# To create model:
model <- lm(Petal.Length ~ Petal.Width, data = iris)
```

Now we can plot out the model

```{r}
plot(model) # 4 plots default-generated, use which = to select what you want

# or
library(ggfortify)
autoplot(model, which = c(1,2))
```



