---
title: "Binomial Formula"
output: rmdformats::robobook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(multicon)
```

# Summary

Binary trial: where only two things can occur, so, *P*(event) = *p* and *P*(non-event) = 1 - *p*. 

- If we have *n* independent, binary trials, with *P*(event) = *p* at every trial, and *n* is fixed; the chance that exactly *k* events occur is 

$$\binom{n}{k}p^x(1-p)^{n-x}$$

- Two things are mutually exclusive when the occurrence of one event prevents the other.

## Box Model

- The box model is a simple way to describe the chance of generating a number. Three things are needed:

    1. The distinct numbers that go in the box ("tickets")
    
    2. The number of each kind of tickets in the box
    
    3. The number of draws from the box. 

- **Sum** of the random draws from a box model with replacement:

$$expected~value~(EV) = number~of~draws \times mean~of~the~box$$

$$standard~error~(SE) = \sqrt{number~of~draws} \times SD~of~box$$

- For the **percentage** of the box (Sum / Sample size):

$$EV~(percentage) = \frac {EV~(sum)}{sample~size} \times 100\%$$
$$SE~(percentage) = \frac {SE~(sum)}{sample~size} \times 100\%$$

## Central Limit Theorem

When drawing (a reasonably large number, like 30) at random with replacement from a box, the probability histogram (in standard units) for the sum (or average) will follow the normal curve. 

# The Box Model

1. What does the "box model" model?

  - The chance of generating a number. 

  - To build the model we need to know:
  
    - the distinct numbers that go in the box ("tickets")
    
    - the number of each kind of tickets in the box
    
    - the number of **draws** from the box. 

2. Suppose we have a box with 10 1's and 5 0's. We take 20 draws (with replacement) and calculate the sum. What is the mean and standard deviation (SD) of the box?

```{r}
# box
box <- c(rep(1,10),rep(0,5)) 
draws <-  20

# mean and SD of box
mean(box)
popsd(box) # sd divides N-1, this divides N
```

3. What is the expected value (EV) and standard error (SE) of the **sum** of 1's in the sample?

```{r}
# EV and SE of sum of the box

# EV = number of draws * mean of box
EVs <- draws*mean(box)

# SE = sqrt(number of draws) * SD of box
SEs <- sqrt(draws)*popsd(box)
c(EVs,SEs)
```

4. What is the expected value (EV) and standard error (SE) of the **proportion** of 1's in the sample?

```{r}
# EV = mean of box
EVp <- mean(box)
# SE = SD of box /sqrt(number of draws)
SEp <- popsd(box)/sqrt(draws)
c(EVp,SEp)
```

5. What is the shape of the sum and proportion of the 1's in the box, with 20 draws?

```{r}
set.seed(10)
data.frame(sum = replicate(20, sum(sample(box, 20, rep = T)))) %>% 
  ggplot(., aes(sum)) + 
  geom_histogram(bins = 10, fill = "lightgrey") + 
  labs(x = "", y = "") + 
  theme_bw()
```

6. If we take a larger number of draws (1000), what shape do we expect the distribution of the sum or proportion to have? What theorem formalizes this result? Demonstrate by simulation.

```{r}
set.seed(10)
data.frame(sum = replicate(20, sum(sample(box, 1000, rep = T)))) %>% 
  ggplot(., aes(sum)) + 
  geom_histogram(bins = 20, fill = "lightgrey") + 
  labs(x = "", y = "") + 
  theme_bw()
```

# Spotify 

## Data Collection (Song List)
Imagine that you have 10 song in a Spotify list. Identify them by 1 to 10.

## Shuffle
In how many different ways could Spotify shuffle this list?

```{r}
factorial(10)
```

## 2 Songs (With Replacement)
Spotify randomly plays 2 songs, with replacement.

1. How many different play lists are there?

```{r}
10*10
```

2. What is the chance of getting a **certain** song twice?

```{r}
1/(10*10)
```

3. What is the chance of getting **any** song twice?

```{r}
1*1/(10)
```

4. Simulate Q3.

```{r}
set.seed(2)
songs <- c(1:10)
replicate(10000, sample(songs, 2, replace = T)) %>%
  which(x = .[1,] == .[2,]) %>%
  length/10000
```

5. What is the chance of not getting a song twice?

6. What is the chance of getting a certain song twice, given it was the 1st song? 
7. Which of the previous events are independent, dependent or mutually exclusive?

8. Would Spotify be likely to draw with replacement?

## 2 Songs (Without Replacement)
Spotify randomly plays 2 songs, without replacement.

1. How many different play lists are there?

```{r}
10*9
```

2. What is the chance of getting your favorite play list?

```{r}
1/(10*9)
```

3. What is the chance of getting a certain song twice? Are these events independent, dependent or mutually exclusive?

```{r}
0
```

4. Simulate this situation.

```{r}
set.seed(2)
songs <- c(1:10)
replicate(10000, sample(songs, 2, replace = F)) %>%
  which(x = .[1,] == .[2,]) %>%
  length/10000
```

## 3 Songs (Without Replacement)
Spotify randomly plays 3 songs, without replacement.

1. If the order matters, how many different playlists are there?

```{r, eval=FALSE}
10*9*8
```

2. If the order doesn't matter, how many different playlists are there?

```{r, eval=FALSE}
choose(10,3) # Which is 10*9*8/factorial(3)
```

3. If order doesn't matter, what is the chance of getting a particular favorite playlist?

4. Assuming each playlist takes 10 minutes, how long would it take to play all possible playlists (in hours)?

## Artist

Consider your playlist in terms of whether the artist is Australian (Y or N). 

1. If Spotify randomly chooses 1 song from your list, what is the chance that the artist is Australian?

For mine, it is actually 0, but let's say if it is 0.4.

2. If Spotify randomly chooses 8 songs from your list (with replacement), what is the chance that just 1 of the artists are Australian?

```{r, eval=FALSE}
choose(8,1)*(0.4)^1*(0.6)^7 # If P(Australian) = 0.4
dbinom(1,8,0.4)
# https://www.statology.org/dbinom-pbinom-qbinom-rbinom-in-r/
```

3. Simulate Q2.

```{r, eval=FALSE}
songs <- c(1,0)  # where "1" = Australian
replicate(1000, sum(sample(songs, 8, replace = T, prob = c(0.4, 0.6)))) %>%
  table(.)/1000
```

4. If Spotify randomly chooses 5 songs from your list (with replacement), what is the chance that no artists are Australian?

## iTunes

How does iTunes shuffle it's playlists?
[Resource](https://www.lifewire.com/is-itunes-shuffle-random-1999249)

Extension: 
[Spotify and Data](https://www.safaribooksonline.com/library/view/strata-data-conference/9781491976326/video314445.html?utm_source=oreilly&utm_medium=newsite&utm_campaign=strata-ny-17-keynote-highlight-music-the-window-into-your-soul)

# Coke 

## Data Collection

Imagine that this is the data for your class:

```{r}
data.frame("Preferred soft drink" = c("Coke", "Pepsi", "Other", "Total"),
           Frequency = c(30, 18, 12, 60))
```


## Domain Knowledge
- According to [past research](http://money.cnn.com/2018/02/20/news/companies/cola-wars-coke-pepsi/index.html), what market share does Coke have?

## Using Binomial
1. Modelling by the Binomial, how many students would you expect to prefer Coke in this class

```{r, eval=FALSE}
n <- 
p <- 
n*p
```

2. Modelling by the Binomial, what is the chance that greater or equal to half the students prefer Coke?

```{r, eval=FALSE}
n <- 
p <- 
1-pbinom(n/2-1,n,p)
```

# iPhones 

## Data Collection

Fill out the following table for your class:

|Preferred mobile phone|Frequency|
|--------------|------------|
|iPhone||
|non-iPhone||
|Total||

```{r}
data.frame("Preferred mobile phone" = c("iPhone", "non-iPhone", "Total"),
           Frequency = c(35, 25, 60))
```


## Domain Knowledge
- According to [past research](https://www.crn.com.au/news/iphones-market-share-begins-to-slide-report-479248), what market share does iPhone have in the US?

## Using Binomial
1. Modelling by the Binomial, how many students would you expect to use an iPhone in this class?

```{r, eval=FALSE}
n <- 
p <- 
n*p
```

2. What are our assumptions?

- The levels of the variable are mutually exclusive (e.g. most students would have either an iPhone *or* another brand, not an iPhone *and* another brand)

3. Modelling by the Binomial, what is the chance that less than or equal of 25% students use iPhones?

```{r, eval=FALSE}
n <- 
p <- 
1 - pbinom(n/4,n,p)
```
